Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-08-03 manuscript.rmd
Minor edits

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS). In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; (Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived from existing data on volumetric water content ($\theta$) vs. $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019]. A Campbell @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:

@@

-- Write about why models were designed this way.

\
\begin{equation}
K_{leaf,z,n} = Ae^{-B\Psi_{soil,~z,n}}
(\#eq:k-soil)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} + error
(\#eq:M1)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,n} StdVPD_{n} + error
(\#eq:M2)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy} + error
(\#eq:M3)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
(\#eq:M4)
\end{equation}

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

Leaf fraction...

For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive Pearson's $\rho$ and *R*^2^ >= 0.1.

@@

\
The model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and abundance Leaf Fraction (Eq. \@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species (*n* = `r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017)

### Realised drought risk integral to assessing drought-vulnerability

Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, that are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015]. Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality. For example, based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].

@@

#### Future Directions for Model Development
\
Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths. Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yr period, which provided large degree of freedom to fit the model. Future studies could better constrain the model with higher resolution growth data such as those from dendrometers. This is especially important for modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing, importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Two of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf fraction is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019]. In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi_{soil}$ vs. VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study; @Chitra-Tarak.2018). Although data needs for parametarizing such models [@Christoffersen.2016tbh; @Sperry.1998] could be greater (e.g. retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g. see @Johnson:2018fad], but which may be potentially harnessed for a community level scale-up.

@@

@@

```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-ksat-profiles} Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively. Only 100 curves are shown for clarity. Note the log scale on horizontal axis."}
knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.pdf"))
```

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors. Mean soil water potential ($Psi_z$) for a given DOY and depth z are shown as lines. Filled areas show extreme drought periods, when $Psi_z$ declined below 2.5 percentile of $Psi_z$ for the given DOY and depth z in any given year within the cenus intervals. Extreme drought periods from multiple years may be overlaid within a panel. Wet season is shown in gray as beginning with DOY "}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
```

```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic concetration ($\\delta^2H$) of soil water by depth measured in March 1997. Data from @Meinzer:1999"}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

@@

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs. $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\circ$^C~ ($gcm^{-3}$).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
```

>>>>

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS). In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived from existing data on volumetric water content ($\theta$) vs. $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019]. A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:

@@

Water is a key limiting resource for tree growth. Vapour pressure deficit (VPD) and species-specific maximum hydraulic conductivity and its decline with water-stress ($K_{leaf}$ vs. $\Psi_{soil}$), are factors known to influence trees water uptake [add references, Chonggang?]. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M4)), in which a species growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as:

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} + error
(\#eq:M1)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~n} StdVPD_{n} + error
(\#eq:M2)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy} + error
(\#eq:M3)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{leaf,~z,~n} LeafCover_{doy}StdVPD_{n} + error
(\#eq:M4)
\end{equation}

We estimated daily value of $K_{leaf,z,n}$ by plugging in $\Psi_{soil,~z,~n}$ in the leaf PLC curves Eq. \@ref(eq:vc-exp) as:

\
\begin{equation}
K_{leaf,~z,~n} = Ae^{-B\Psi_{soil,~z,~n}}
(\#eq:k-soil)
\end{equation}

We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015; see below).

@@

To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) that are likely to be fully exposed to the sun. We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (Do I need to add SI figure?). We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).

As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) data we collected from the Eddy Flux Tower and VPD (kPa; $\bar{GPP} = m_2VPD^2 + m_1VPD + m_0 + error = StdVPD$; Do I need to add SI Fig?) and used standardised $StdVPD$ (range 0-1) in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

To explore the effect of species seasonality in leaf cover on growth, we made an informed assumption of  species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless). For evergreen species we assumed $LeafCover$ = 1 throughtout the year, whereas for species with all other decisuous classes, we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300), given high forest leaf area index for this period (data not shown). We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). For non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (\@ref(fig:leaf-cover)).

We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section.

For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp and model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD), only for species with positive Pearson's $\rho$ and *R*^2^ >= 0.1.

@@

\
The model with VPD and without Leaf Cover (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and Leaf Cover (Eq. \@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with Leaf Cover had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Cover (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$). See SI for data collection and estimation of these variables.

@@

We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$ 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and $time$ is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$ and, finally, species mortality anomalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and jittered vertically around the named depths (ticks) for clarity. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species (*n* = `r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017).

### Realised drought risk integral to assessing drought-vulnerability
\
Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, that are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015]. Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, (median; `r round(as.numeric(psi.1.7.min$q2.5), 2)`, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality. For example, based on a year-long study at BCI, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].

@@

### Future Directions for Model Development
\
Our model estimated ERD as a single value, unlike the dynamic ERD modelled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths. Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25-yr period, which provided large degree of freedom to fit the model. Future studies could better constrain the model with higher resolution growth data such as those from dendrometers. This is especially important for modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing, importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Two of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\delta^2H$, Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explaine growth, but also acknowledge that our estimate of seasonality of leaf cover is a tentative estimate based on leaf fall data and assumption of leaf presence during the wet season and omits inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019]. In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi_{soil}$ vs. VWC), hydraulic conductivity ($K_{soil}$) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018). Although data needs for parametarizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g. retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g. see @Johnson:2018fad], but which may be potentially harnessed for a community level scale-up.

@@

Yet To be written.

@@

```{r ksat-profiles, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ksat-profiles} Generated 5000 curves of soil hydraulic conductivity variation by depth based on Latin Hypercube Sampling within 95% CI of observed data at depths of 12.5 cm and 60 cm [@Kinner:2004], assuming a linear decline between those depths, and values before 12.5 and after 60 cm as that of 12.5 cm and 60 cm, respectively. Only 100 curves are shown for clarity. Note the log scale on horizontal axis.")}
knitr::include_graphics(file.path("figures/Bootstrapped Ksat profiles within observed 95CI_random_two_hundred.pdf"))
```

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts} Extreme soil droughts by census interval (panels) at three representative depths: 0.5, 1 and 1.7 m, depicted in different colors. Mean soil water potential ($Psi_{soil,~z}$) for a given DOY and depth z are shown as lines. Filled areas show extreme drought periods, when $Psi_{soil,~z}$ declined below 2.5 percentile of $Psi_{soil,~z}$ for the given DOY and depth z in any given year within the cenus intervals. Extreme drought periods from multiple years may be overlaid within a panel. Wet season is shown in gray as beginning with DOY."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_depths_in_inverse_model.jpeg"))
```

```{r iso-data, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-iso-data}Stable isotopic concetration ($\\delta^2H$) of soil water by depth measured in March 1997. Data from @Meinzer:1999."}
knitr::include_graphics(file.path("figures/UDI_confidence/Meinzer_etal_1999_Depth_vs_soil_deltaD.jpeg"))
```

@@

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs. $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
```
